---
title: "Dimensionality Reduction"
author: "Brylle Joshua Logroño"
date: "2023-08-02"
output: html_document
---

#### This document contains dimensionality reduction contents.

## Time Complexity of PCA, t-SNE, and UMAP

### PCA
PCA (Principal Component Analysis) is a widely used technique for dimensionality reduction. The time complexity of PCA depends on the number of data-points (n) and the number of features (p). In the worst-case scenario (based on the definition), the time complexity of PCA is O(p^2n + p^3), where n is the number of data-points and p is the number of features.

The computation of the covariance matrix contributes O(p^2n) to the overall complexity, while the eigen-value decomposition adds O(p^3). In R, using the `prcomp` function with scaling to implement PCA as SVD (Singular Value Decomposition) of the Dataset matrix results in a time complexity of O(n^2p + p^3).

For reference and more information:
- [StackOverflow - Complexity of PCA](https://stackoverflow.com/questions/20507646/how-is-the-complexity-of-pca-ominp3-n3)
- [Medium - Computational Complexity of PCA](https://alekhyo.medium.com/computational-complexity-of-pca-4cb61143b7e5)
- [SVD Implementation to Speed up PCA](https://www.slideshare.net/YounesCharfaoui/principal-component-analysis-code-and-time-complexity-127431697)

### t-SNE
t-SNE (t-distributed Stochastic Neighbor Embedding) is another dimensionality reduction technique. Its time complexity is O(n^2), where n is the number of data-points.

For reference and more information:
- [t-SNE Paper (arXiv)](https://arxiv.org/pdf/1512.01655.pdf)

### UMAP
UMAP (Uniform Manifold Approximation and Projection) is a popular dimensionality reduction method. Its time complexity varies based on the dimension of the target reduced space (d) and the number of data-points (n). In the average case, the complexity is O(d * n^1.14), while in the worst-case scenario, it is O(n^2).

For reference and more information:
- [UMAP GitHub Issue](https://github.com/lmcinnes/umap/issues/8)
- [ResearchGate - UMAP Publication](https://www.researchgate.net/publication/323141395)


```{r include-functions, echo = FALSE}

# Function to extract the timestamp from the kmer files
extract_time <- function(string) {
  parts <- strsplit(string, "_")[[1]]
  as.numeric(gsub(".csv", "", parts[3]))
}

# Function to search and read the CSV file
read_kmer_csv <- function(data_path, k) {
  print("Reading CSV file...")
  # Get the list of files matching the pattern
  file_pattern <- paste0("kmer_", k, "_", ".*\\.csv$")
  file_list <- list.files(
    path = data_path, pattern = file_pattern,
    full.names = FALSE
  )
  
  # Check if any files are found
  if (length(file_list) == 0) {
    message("No files found for k = ", k)
    return(NULL)
  }
  
  # Sort the strings based on the timestamp in descending order
  sorted_strings <- file_list[order(sapply(file_list, extract_time),
                                    decreasing = TRUE
  )]
  
  df <- read.csv(paste(data_path, sorted_strings[1], sep = "/"))
  return(df)
}

# Function for saving 2D plots as PNG and HTML
save_plot <- function(method, results_path, k, p, is_3d = FALSE) {
  print("Saving plot...")
  # File name for saving
  filename <- paste0(method, "-", k, ".png")
  
  # Note: 3D plots are plot_ly objects, 2D plots are ggplot objects.
  if (is_3d) {
    # Save plot_ly obj. as PNG
    # save_image(p, paste(results_path, filename, sep = "/"))
    
    # Convert ggplot object to ggplotly
    p <- ggplotly(p) 
    # Save as RData
    save(p, file = file.path(results_path, paste0(method, "-", k, ".RData")))
    
  } else {
    # Save as PNG
    ggsave(filename, p, results_path,
           device = "png", width = 5, height = 5,
           dpi = 300, bg = "white"
    )
    # Convert ggplot object to ggplotly
    p <- ggplotly(p) 
    # Save as RData
    save(p, file = file.path(results_path, paste0(method, "-", k, ".RData")))
  }
  
  # Save as HTML
  html_file <- paste0(results_path, "/", method, "-", k, ".html")
  saveWidget2(widget = p, file = html_file)
}

# Function for pre-processing and scaling of data
pre_process <- function(df) {
  print("Pre-processing and scaling data...")
  # Extract year from date column
  
  # Determine the columns to use (drop metadata, retain k-mers)
  slice_col <- which(colnames(df) == "strain")
  x <- df[, 2:(slice_col - 1)]
  # color <- df[[factor1]]
  
  # Check for columns that have zero variance for PCA
  non_zero_var_cols <- apply(x, 2, var) > 0
  x <- x[, non_zero_var_cols]
  
  if (ncol(x) < 2) {
    stop("Insufficient columns with non-zero variance for PCA.")
  }
  
  # Scale data
  x <- scale(x)
  
  # (This is needed for labeling of points in 3D plots)
  df$year <- format(as.Date(df$date), "%Y")
  
  return(x)
}

# Function to execute before main dim-reduce codes
pre_reduce <- function(results_path, kmers, k, filter1_factor, 
                       filter1_values, filter2_factor, filter2_values) {
  # Check if the directory already exists
  if (!dir.exists(results_path)) {
    # Create the directory if it doesn't exist
    dir.create(results_path, recursive = TRUE)
  }
  # Process kmers dataframe
  df <- kmers
  df$year <- format(as.Date(df$date), "%Y")
  
  # Making filter values optional
  # if(missing(filter1_factor) && missing(filter1_values) &&
  #    missing(filter2_factor) && missing(filter2_values)){
  #   
  # } else if(missing(filter2_factor) && missing(filter2_values)){
  #   
  #   df <- filter(df, df[[filter1_factor]] %in% filter1_values)
  #   
  # } else if(missing(filter1_factor) && missing(filter1_values)){
  #   
  #   df <- filter(df, df[[filter2_factor]] %in% filter2_values)
  #   
  # } else {
  #   df <- filter(df, df[[filter1_factor]] %in% filter1_values)
  #   df <- filter(df, df[[filter2_factor]] %in% filter2_values)
  # }
  
  # Pre-process the data
  x <- pre_process(df)
  return(list(df = df, x = x))
}

# Function that performs PCA
pca_fn <- function(x) {
  print("Performing PCA...")
  pca_df <- prcomp(x, center = TRUE)
  return(pca_df)
}

# Function for 2D PCA plot
pca_2d <- function(pca_df, df, k, color, shape, results_path) {
  color <- df[[color]]
  shape <- df[[shape]]
  print("Generating 2D PCA plot...")
  p <- autoplot(pca_df, data = df) +
    geom_point(aes(color = color, shape = shape, text = paste(
      "Identifier: ", df$gisaid_epi_isl, "\n",
      "Variant: ", df$variant, "\n",
      "Sex: ", df$sex, "\n",
      "Division Exposure: ", df$division_exposure, "\n",
      "Year: ", format(as.Date(df$date), "%Y"), "\n",
      "Strain: ", df$strain, "\n",
      "Pangolin Lineage: ", df$pangolin_lineage
    ))) +
    scale_color_brewer(palette = "Set1")
  # Save plot as PNG and HTML
  save_plot("2d-pca", results_path, k, p)
}

# Function for 3D PCA plot
pca_3d <- function(pca_df, df, k, color, shape, results_path) {
  print("Generating 3D PCA plot...")
  pc <- as.data.frame(pca_df$x[, 1:3])
  
  # Use plot_ly for 3D visualization
  p <- plot_ly(pc,
               x = ~PC1, y = ~PC2, z = ~PC3, type = "scatter3d",
               mode = "markers", color = df[[color]], shape = df[[shape]],
               text = paste(
                 "Identifier: ", df$gisaid_epi_isl, "\n",
                 "Variant: ", df$variant, "\n",
                 "Sex: ", df$sex, "\n",
                 "Division Exposure: ", df$division_exposure, "\n",
                 "Year: ", format(as.Date(df$date), "%Y"), "\n",
                 "Strain: ", df$strain, "\n",
                 "Pangolin Lineage: ", df$pangolin_lineage
               )
  )
  
  # Save plot as PNG and HTML
  save_plot("3d-pca", results_path, k, p, is_3d = TRUE)
}

# Function that generates scree plot from PCA results
screeplot <- function(pca_df, k, results_path) {
  print("Generating scree plot...")
  p <- fviz_eig(pca_df,
                xlab = "Number of Principal Components"
  )
  
  # Save plot as PNG and HTML
  save_plot("screeplot", results_path, k, p)
}

# Function that generates factor loadings of first
# n_components Principal Components
factor_loadings <- function(pca_df, x, k, n_components, results_path) {
  print(paste("Generating factor loadings plot of first", n_components,
              "PCs...",
              sep = " "
  ))
  # Extract factor loadings
  loadings <- pca_df$rotation
  
  # Plot bar plots for factor loadings of n principal components
  for (i in 1:n_components) {
    # Create a data frame for the factor loadings
    loadings_df <- data.frame(
      variable = colnames(x),
      loading = loadings[, i]
    )
    
    # Create a bar plot using ggplot2
    p <- ggplot(loadings_df, aes(x = variable, y = loading)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(
        title = paste("Principal Component", i),
        x = "Variables", y = "Factor Loadings"
      )
    
    # Save plot as PNG and HTML
    save_plot(paste("loadings", i, sep = "-"), results_path, k, p)
  }
}

# Function that generates graph of individuals from PCA results
indiv <- function(pca_df, k, results_path) {
  print("Generating graph of individuals...")
  p <- fviz_pca_ind(pca_df,
                    col.ind = "cos2", # Color by the quality of representation
                    gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                    repel = TRUE, # Avoid text overlapping
                    xlab = "PC1",
                    ylab = "PC2",
  )
  
  # Save plot as PNG and HTML
  save_plot("indivgraph", results_path, k, p)
}

# Function that generates graph of variables from PCA results
vars <- function(pca_df, k, results_path) {
  print("Generating graph of variables...")
  p <- fviz_pca_var(pca_df,
                    col.var = "contrib", # Color by contributions to the PC
                    gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                    repel = TRUE, # Avoid text overlapping
                    xlab = "PC1",
                    ylab = "PC2"
  )
  
  # Save plot as PNG and HTML
  save_plot("vargraph", results_path, k, p)
}

# Function that generates biplot from PCA results
biplot <- function(pca_df, k, results_path) {
  print("Generating biplot...")
  # # [Old method] Create biplot of individuals and variables
  p <- fviz_pca_biplot(pca_df,
                       col.var = "#2E9FDF", # Variables color
                       col.ind = "#696969", # Individuals color
                       addEllipses = TRUE,
                       xlab = "PC1",
                       ylab = "PC2"
  )
  
  # # Create biplot of individuals and variables (using ggbiplot)
  # p <- ggbiplot(pca_df,
  #   obs.scale = 1, var.scale = 1,
  #   groups = color, ellipse = TRUE, circle = TRUE
  # ) +
  #   scale_color_discrete(name = "") +
  #   theme(legend.direction = "horizontal", legend.position = "top")
  
  # Save plot as PNG and HTML
  save_plot("biplot", results_path, k, p)
}

# Function that performs t-SNE (Rtsne library)
rtsne_fn <- function(pca_results, tsne_dims, tsne_perplexity, tsne_max_iter, 
                     tsne_seed) {
  print("Performing t-SNE...")
  set.seed(tsne_seed)
  tsne_df <- Rtsne(pca_results,
                   dims = tsne_dims, perplexity = tsne_perplexity,
                   max_iter = tsne_max_iter, check_duplicate = FALSE,
                   pca = FALSE
  )
  return(tsne_df)
}

# Function that includes visualization for each t-SNE iteration
ecb <- function(x) {
  epoc_df <- data.frame(x, color = factor1, shape = factor2)
  
  plt <- ggplot(epoc_df, aes(
    x = X1, y = X2,
    label = color, color = color, shape = shape
  )) +
    geom_text()
  
  print(plt)
}

# Function that performs t-SNE (tsne library)
tsne_fn <- function(pca_results, tsne_dims, tsne_initial_dims, tsne_perplexity, 
                    tsne_max_iter, tsne_seed) {
  print("Performing t-SNE...")
  set.seed(tsne_seed)
  if (tsne_dims == 2) {
    tsne_df <- tsne(pca_results,
                    k = tsne_dims,
                    initial_dims = tsne_initial_dims,
                    perplexity = tsne_perplexity,
                    max_iter = tsne_max_iter,
                    # epoch_callback = ecb
    )
  } else {
    tsne_df <- tsne(pca_results,
                    k = tsne_dims,
                    initial_dims = tsne_initial_dims,
                    perplexity = tsne_perplexity,
                    max_iter = tsne_max_iter
    )
  }
  
  return(tsne_df)
}

# Function for 2D t-SNE plot
tsne_2d <- function(tsne_df, df, color, shape, k, is_tsne, results_path) {
  print("Generating 2D t-SNE plot...")
  if (is_tsne) {
    tsne_df <- data.frame(X1 = tsne_df[, 1], X2 = tsne_df[, 2], 
                          color = df[[color]], shape = df[[shape]])
  } else {
    tsne_df <- data.frame(X1 = tsne_df$Y[, 1], X2 = tsne_df$Y[, 2], 
                          color = df[[color]], shape = df[[shape]])
  }
  # Create ggplot object
  p <- ggplot(tsne_df, aes(x = X1, y = X2, color = color, shape = shape)) +
    geom_point(aes(text = paste(
      "Identifier: ", df$gisaid_epi_isl, "\n",
      "Variant: ", df$variant, "\n",
      "Sex: ", df$sex, "\n",
      "Division Exposure: ", df$division_exposure, "\n",
      "Year: ", format(as.Date(df$date), "%Y"), "\n",
      "Strain: ", df$strain, "\n",
      "Pangolin Lineage: ", df$pangolin_lineage
    ))) +
    xlab("TSNE-2D-1") +
    ylab("TSNE-2D-2") +
    scale_color_brewer(palette = "Set1")
  
  # Save plot as PNG and HTML
  save_plot("2d-tsne", results_path, k, p)
}

# Function for 3D t-SNE plot
tsne_3d <- function(tsne_df, df, color, shape, k, results_path) {
  print("Generating 3D t-SNE plot...")
  final <- cbind(data.frame(tsne_df), df[[color]], df[[shape]])
  p <- plot_ly(final,
               x = ~X1, y = ~X2, z = ~X3, type = "scatter3d", mode = "markers",
               color = df[[color]], shape = ~shape,
               text = paste(
                 "Identifier: ", df$gisaid_epi_isl, "<br>",
                 "Variant: ", df$variant, "<br>",
                 "Sex: ", df$sex, "<br>",
                 "Division Exposure: ", df$division_exposure, "<br>",
                 "Year: ", format(as.Date(df$date), "%Y"), "<br>",
                 "Strain: ", df$strain, "<br>",
                 "Pangolin Lineage: ", df$pangolin_lineage
               )
  )
  
  # Save plot as PNG and HTML
  save_plot("3d-tsne", results_path, k, p, is_3d = TRUE)
}

# Function that performs UMAP
umap_fn <- function(x, umap_dims, umap_n_neighbors, umap_metric, 
                    umap_min_dist, umap_seed) {
  print("Performing UMAP...")
  umap_df <- umap(x,
                  n_components = umap_dims, n_neighbors = umap_n_neighbors,
                  metric = umap_metric, min_dist = umap_min_dist,
                  random_state = umap_seed
  )
  return(umap_df)
}

# Function for 2D UMAP plot
umap_2d <- function(umap_df, df, color, shape, k, results_path) {
  print("Generating 2D UMAP plot...")
  emb <- umap_df$layout
  
  x_o <- emb[, 1]
  y_o <- emb[, 2]
  
  # Create ggplot object
  p <- ggplot(
    data = as.data.frame(emb),
    aes(x = x_o, y = y_o, color = df[[color]], shape = df[[shape]])
  ) +
    geom_point(aes(text = paste(
      "Identifier: ", df$gisaid_epi_isl, "\n",
      "Variant: ", df$variant, "\n",
      "Sex: ", df$sex, "\n",
      "Division Exposure: ", df$division_exposure, "\n",
      "Year: ", format(as.Date(df$date), "%Y"), "\n",
      "Strain: ", df$strain, "\n",
      "Pangolin Lineage: ", df$pangolin_lineage
    ))) +
    xlab("UMAP_1") +
    ylab("UMAP_2") +
    labs(shape = "shape", color="color") +
    scale_color_brewer(palette = "Set1")
  
  # Save plot as PNG and HTML
  save_plot("2d-umap", results_path, k, p)
}

# Function for 3D UMAP plot
umap_3d <- function(umap_df, df, color, shape, k, results_path) {
  print("Generating 3D UMAP plot...")
  final <- cbind(data.frame(umap_df[["layout"]]), color, shape)
  p <- plot_ly(final,
               x = ~X1, y = ~X2, z = ~X3, type = "scatter3d", mode = "markers",
               color = df[[color]], shape = ~shape,
               text = paste(
                 "Identifier: ", df$gisaid_epi_isl, "<br>",
                 "Variant: ", df$variant, "<br>",
                 "Sex: ", df$sex, "<br>",
                 "Division Exposure: ", df$division_exposure, "<br>",
                 "Year: ", format(as.Date(df$date), "%Y"), "<br>",
                 "Strain: ", df$strain, "<br>",
                 "Pangolin Lineage: ", df$pangolin_lineage
               )
  )
  
  p <- p %>% add_markers()
  p <- p %>% layout(scene = list(
    xaxis = list(title = "0"),
    yaxis = list(title = "1"),
    zaxis = list(title = "2")
  ))
  
  # Save plot as PNG and HTML
  save_plot("3d-umap", results_path, k, p, is_3d = TRUE)
}

```

### Dimensionality Reduction Code
These are the needed parameters:
k
kmers
results_path
tsne_seed
tsne_perplexity
tsne_max_iter
tsne_initial_dims
umap_seed
umap_n_neighbors
umap_metric
umap_min_dist
color
shape
include_plots
filter1_factor
filter1_values
filter2_factor
filter2_values

## Dimensionality Reduction Methods
### Principal Component Analysis (PCA)
PCA is a popular method for dimensionality reduction that transforms high-dimensional data into a lower-dimensional space. It identifies the principal components, which are linear combinations of the original features that capture the most variance in the data. In this code, PCA is used to reduce the dimensionality of the scaled data (x). The results are plotted in 2D and 3D PCA plots, screeplot, factor loadings plot, and many other more.

```{r pca}
pre_reduce_res <- pre_reduce(results_path, kmers, k, filter1_factor, 
                               filter1_values, filter2_factor, filter2_values)
  
df <- pre_reduce_res$df                # df is the original dataset
x <- pre_reduce_res$x                  # x is the scaled data

# Perform PCA
pca_df <- pca_fn(x)
```

### t-Distributed Stochastic Neighbor Embedding (t-SNE)
t-SNE is a nonlinear dimensionality reduction technique that is particularly useful for visualizing high-dimensional data in a lower-dimensional space while preserving local structures. In this code, t-SNE is performed using the tsne or Rtsne library (based on user selection) on the PCA results. The function generates 2D and 3D t-SNE plots, which represent the data points in a way that maintains the similarity between nearby points.

```{r tsne}
# Perform t-SNE via 'tsne' library using PCA results (in 3 dimensions)
# # Note: Uncomment the next two line to use tsne; otherwise, comment them
is_tsne <- TRUE
tsne_df <- tsne_fn(pca_df$x, 3, tsne_initial_dims, tsne_perplexity,
                   tsne_max_iter, tsne_seed)

# Perform t-SNE via 'Rtsne' library using PCA results (in 3 dimensions)
# # Note: Uncomment the next two lines to use Rtsne; otherwise, comment them
# is_tsne <- FALSE
# tsne_df <- rtsne_fn(pca_df$x, 3, tsne_perplexity, tsne_max_iter, tsne_seed)
```

### Uniform Manifold Approximation and Projection (UMAP)
UMAP is another nonlinear dimensionality reduction method that seeks to preserve both local and global structures of the data. It is known for its ability to scale to large datasets. In this code, UMAP is used to project the scaled data (x) into a 3D space. The function generates 2D and 3D UMAP plots, which provide a reduced representation of the data while preserving the relationships between data points.

```{r umap}
# Perform UMAP (in 3 dimensions)
umap_df <- umap_fn(x, 3, umap_n_neighbors, umap_metric, 
                   umap_min_dist, umap_seed)
```

### 2D PCA plot:

- This plot represents the data points projected onto the first two principal components obtained from PCA.
- It helps visualize the overall distribution of the data in a 2D space while retaining the most significant variability in the data.
- The plot may reveal clusters or patterns in the data.

```{r 2dpca}
# Generate 2D PCA plot
pca_2d(pca_df, df, k, color, shape, results_path)
```

### 3D PCA plot (pca_3d):
- Similar to the 2D PCA plot, but in a 3D space.
- It provides additional depth to visualize the data points and explore the variation in the first three principal components.
- It allows for the identification of clusters or patterns that might not be apparent in the 2D projection.

```{r 3dpca}
# Generate 3D PCA plot
pca_3d(pca_df, df, k, color, shape, results_path)
```

### Screeplot:
- A screeplot is a line plot that shows the variance explained by each principal component.
- It helps identify the number of principal components that capture most of the variance in the data.
- The "elbow point" in the plot is often used to determine the appropriate number of principal components to retain.

```{r scree}
# Generate screeplot
screeplot(pca_df, k, results_path)
```


### Factor Loadings plot (factor_loadings):
- This plot shows the loadings of variables on the first three principal components.
- It helps in understanding which variables contribute most to the principal components.
- Variables with high absolute loadings contribute significantly to the components' variation.

```{r loadings}
# Generate factor loadings plot of first 3 principal components
factor_loadings(pca_df, x, k, 3, results_path)
```

### Graph of Individuals:
- Also known as a "score plot" or "individual plot," it displays the data points in the PCA space based on their scores on the principal components.
- It helps to visualize how individual samples are distributed in the reduced PCA space.
- Clusters of points close together may indicate similar samples in terms of the measured variables.

```{r indiv}
# Generate graph of individuals
indiv(pca_df, k, results_path)
```

### Graph of Variables (vars):
- This plot shows the projection of the original variables onto the first two principal components.
- It helps to understand how variables are related to each other and how they contribute to the PCA components.
- Variables close to each other in the plot are positively correlated, while variables far apart are negatively correlated.

```{r vars}
# Generate graph of variables
vars(pca_df, k, results_path)
```

### Biplot:
- A biplot combines the individual and variable plots into a single plot.
- It helps to visualize both the data points and the variables in the same reduced-dimensional space.
- This enables understanding the relationships between variables and individuals.

```{r biplot}
# Generate biplot
biplot(pca_df, k, results_path)
```

### 2D t-SNE plot (tsne_2d):
- This plot represents the data points in a 2D space, using the t-distributed Stochastic Neighbor Embedding (t-SNE) algorithm.
- t-SNE is useful for visualizing high-dimensional data and preserving local structures.
- Clusters of points that appear close together in this plot are likely to be similar in the original high-dimensional space.

```{r 2dtsne}
# Generate 2D t-SNE plot
tsne_2d(tsne_df, df, color, shape, k, is_tsne, results_path)
```

### 3D t-SNE plot
- Similar to the 2D t-SNE plot, but in a 3D space, providing additional perspectives on the data distribution.
- It can help identify more complex and intricate structures within the data.

```{r 3dtsne}
# # Note: Uncomment the succeeding line for tsne;
# # otherwise, comment it
# Generate 3D t-SNE plot
tsne_3d(tsne_df, df, color, shape, k, results_path)

# # Note: Uncomment the succeeding line for Rtsne;
# # otherwise, comment it
# tsne_3d(tsne_df$Y, df, color, shape, k, results_path)
```

### 2D UMAP plot
- This plot represents the data points in a 2D space using the Uniform Manifold Approximation and Projection (UMAP) algorithm.
- UMAP is another dimensionality reduction technique similar to t-SNE, but often computationally more efficient.
- It can reveal different structures or patterns in the data compared to t-SNE.

```{r 2dumap}
# Generate 2D UMAP plot
umap_2d(umap_df, df, color, shape, k, results_path)
```


### 3D UMAP plot (umap_3d):
- Similar to the 2D UMAP plot, but in a 3D space, providing additional perspectives on the data distribution.
- It allows for exploring the data points' distribution and potential clusters in a more complex space.

```{r 3dumap}
# Generate 3D UMAP plot
umap_3d(umap_df, df, color, shape, k, results_path)
```